Nuclear Waste Management and Disposal

### 2017-02-27 paper.tex
ready version 2 modulo typos

\date{\today}%
\begin{abstract}
%    As nuclear reactors around the world keep operating with no disposal method available for highly
%    radioactive nuclear waste, how we manage nuclear waste is becoming more important. In this work,
%    we examine a few scientific and technical related to nuclear waste management and disposal.
%    The principles of radioactivity and nuclear fission helps us understand the source and properties
%    of different types of nuclear waste. Focusing on the highly radioactive spent fuel, the rest
%    of this work outlines the advantages and drawbacks of different spent fuel management and disposal
%    methods. We study the methods used for performance assessments for Yucca Mountain Nuclear Waste Repository,
%    especially in the context of volcanic disruption. Finally, we discuss some possible options for 
%    nuclear waste management and disposal after
%    the termination of the Yucca Mountain project.\\
%    Word count: 6592\\
\end{abstract}
\maketitle
\tableofcontents
\newpage
\section{Introduction}
Nuclear power continues to play an important role in global energy supply.
The International Atomic Energy Agency (IAEA) estimates that nuclear power
consumption will increase by $40\%$ by 2030.\cite{iaea12}
Whereas the ``front end'' of the nuclear fuel cycle, such as reactor physics and power output, draws
the most attention from the public, the ``back end'' of the fuel cycle, i.e. waste management and
disposal, deserves equal, if not more attention.
Currently, the IAEA estimates that there are over 10,000 metric tons of highly radioactive nuclear waste 
generated worldwide annually,
most of which is spent nuclear fuel from nuclear reactors.\cite{iaea08, r12}
The world will have over $457,000$ metric tons of highly radioactive waste stock by the end of 2020.\cite{r12}
Highly radioactive waste remains hazardous for up to $10^6$ years after it is produced and thus
requires extremely careful disposal methods.
Nevertheless, there is currently no permanent repository in the world to dispose of these highly radioactive
waste.\footnote{The Waste Isolation Pilot Plant (WIPP) in New Mexico is designed to be a permanent radioactive
    waste repository, but only
for waste classified as Transuranic, which, by the Nuclear Regulatory Commission standard, does not have particularly
high radioactivity but only long half-life (See Sec.~\ref{sec:tru}). In particular, WIPP does not accept spent nuclear
fuel.}

@@

Evidently, nuclear waste management and disposal is becoming a increasingly urgent matter globally.
In light of the high stake,
the uncertainty of the stability of a geological repository over a extremely long timescale has evolved into 
a topic not only of scientific contention but also of policy debate.
How well is the predicative power of science? How much trust should we put into the uncertainty estimates
(i.e. \red{how well can science predict its predicative power})?

@@

\section{Radioactivity and Its Biological Effects}
\label{sec:phys}
    Here we detail the nuclear physics underlying most of the discussions
    in this work. We will apply these results in our discussion of nuclear wastes.
    \subsection{Radioactivity}
    Radioactive is the process in which most of the known detrimental radiations are generated.
    The derivations and in this section follows chapter $6$ in
    Krane's text \cite{k88} and chapter $1$ in Lilley's text \cite{l01}.
        \subsubsection{Quantum theory of radioactive decays}
        The law of radioactive decay is a direct consequence of it being a quantum-mechanical
        effect.
        We can approximate the nucleus as a particle situated in a potential well with potential given by
        \begin{equation}
            \label{eq:vp}
            \tilde{V}(t) = V + V_p,
        \end{equation}
        where $V$ is the time-independent nucleus potential that gives rise to eigenstates $\ket{\psi_1}$
        and $\ket{\psi_2}$ with $\ket{\psi_2}$ having a lower (expectation value of) energy; $V_p$ is a small
        perturbation that causes the decay.
        Suppose the particle is in a state given by
        \begin{equation}
            \label{eq:state}
            \ket{\psi(t)} = c_1(t)\ket{\psi_1} + c_2(t)\ket{\psi_2},
        \end{equation}
        where $c_1,c_2$ are the coefficients of superposition.
        A decay from $\ket{\psi_1}$ to $\ket{\psi_2}$ is equivalent to
        having $|c_2(t)|^2\gg |c_1(t)|^2$ as time increases. Since we know that time-independent potentials
        only give rise to time-independent states, the fact that Eq.~\ref{eq:state} has time dependence requires
        our perturbation $V_p$ in Eq.~\ref{eq:vp} to be time dependent.

@@

        Specific activity, on the other hand, gives the activity per mass of a sample
        \begin{equation}
        \mathcal{A}_s = \frac{\mathcal{A}(0)}{{m_{sample}}}.
        \end{equation}
        Specific activity can characterize the activity level of a certain isotope or compound.
        The difference between
        activity and specific activity resembles the difference between heat capacity and specific heat
        capacity.
        The radioactive isotope used for Computer Tomography (CT) scan has a specific activity
        of $\sim 10^{8}$ Bq/g. See Table~\ref{tab:hl} for the half-lives and specific activities
        of some isotopes present in highly radioactive nuclear waste.
        \begin{table}
            \centering
            \caption{Specific activity of isotopes commonly present in spent nuclear fuel \cite{toi}}
            \begin{ruledtabular}
                \begin{tabular}{l l l}
                Isotope & Half-life (Years) & Specific Activity (Ci/g)\\
                \hline
                Cesium-137 & $37$ & $88$\\
                Radium-226 & $1.6\times 10^3$ & $1$\\
                Plutonium-239 & $2.4\times 10^4$ & $0.063$\\
                Uranium-233 & $1.6\times 10^6$ & $9.3\times 10^{-3}$\\
                Thorium-232 & $1.4 \times 10^{10}$ & $1.1\times 10^{-7}$\\
            \end{tabular}
            \label{tab:hl}
            \end{ruledtabular}
        \end{table}
        \subsubsection{Types of radiation}
        There are three main types of radioactive decay processes that generate radiations in nuclear waste: 
        Alpha decay, which generates Alpha particles (helium-4 nuclei), Beta decay which generates Beta
        particles (electrons) and Gamma decay which generates gamma rays.
        The type of radiation (particle beam) generated
        have different properties and deserve separate trearments when we discuss health effects and shielding.
        \paragraph{Alpha decay}
        An alpha particle is a helium-4 atom ${}^4_2$He. In an alpha decay, a nucleus emits an alpha
        particle and decays into a different element. The emission of an alpha particle is
        particularly favorable because it does not change the parity of the nucleus and has a particularly
        high binding energy. The process can be represented by
        \begin{equation}
            {}^A_Z X_N \rightarrow {}^{A-4}_{Z-2}X'_{N-2} + \alpha.
        \end{equation}

@@

        \subsubsection{Daughter nuclei and subsequent decay}
        The resultant nuclei (called ``daughter products'') from a radioactive decay are oftern radioactive
        themselves and their decay rate can be larger or smaller than their parents. Consider a decay chain
        $A\rightarrow B \rightarrow C \rightarrow ...$ with decay constants $\lambda_A, \lambda_B,...$. Suppose
        we start with a sample of $N_0$ $A$ nuclei. The change of the number of nuclei $B$ increases as $A$
        decays and decreases s $B$ decay. Hence we have
        \begin{equation}
            \frac{dN_B}{dt} = -\lambda_B N_B + \lambda_A N_A.
        \end{equation}
        Multiply both sides by $e^{\lambda_B t}$ and substitute in $N_A=N_0 e^{-\lambda_A} t$ to get
        \begin{align*}
            \frac{dN_B}{dt}e^{\lambda_B t} + \lambda_B N_B e^{\lambda_B t} &= \lambda_A N_A e^{\lambda_B},\nonumber\\
            \frac{d}{dt} (N_B e^{\lambda_B t}) &= \lambda_A N_0 e^{(\lambda_B-\lambda_A)t}, \nonumber\\
            N_B e^{\lambda_B t} &= \lambda_A N_0 \int\limits_0^t e^{(\lambda_B - \lambda_A)u} du \nonumber\\
                                &= \frac{\lambda_A}{\lambda_B-\lambda_A}N_0 e^{(\lambda_B - \lambda_A)t}
        \end{align*}
        Hence we end up with the number of $B$ nuclei as a function of time
        \begin{equation}
            N_B(t)=\frac{\lambda_A}{\lambda_B-\lambda_A}N_0(e^{-\lambda_A t} - e^{-\lambda_B t}).
            \label{eq:daughter}
        \end{equation}
        Similar derivations can be done for the rest of the isotopes in the decay chain by considering
        one decay in a chain at a time. If the daughter isotope has a longer half-life than the parent
        ($\lambda_A \gg \lambda_B$), for small $t$, Eq.~\ref{eq:daughter} becomes
        $N_B(t)\sim N_0 e^{-\lambda_A t}$. After a long time,
        Eq.~\ref{eq:daughter} becomes $N_B(t) \sim N_0 e^{-\lambda_B t}$; the decay
        chain can be treated as the fast decay of $A$ into $B$, and then the slow decay of $B$
        into $C$.

        On the other hand, if the parent is longer lived
        ($\lambda_B \gg \lambda_A)$, after a long time, Eq.~\ref{eq:daughter} can be written as
        \begin{equation}
            N_B(t)=\frac{\lambda_A}{\lambda_B}N_0 e^{-\lambda_A t}=\frac{\lambda_A}{\lambda_B}N_A(t).
        \end{equation}
        Hence the number of nuclei $A$ and nuclei $B$ are in equilibrium determined by their decay rates, with
        $A$ dominating the nuclei count. Taking the limit as $\lambda_A \rightarrow 0$, $N_B(t)\rightarrow 0$.
        The decay is essentially the decay of $A$ directly into $C$. Hence, if we have a chain of reaction
        where each daughter nuclei has a half-life much shorter than that of its parent, then 
        isotope $A$ effectively decays into the end of the decay chain and releases a lot more radiations than
        in a single decay. Moreover, before the decay chain reaches its equilibrium, the \red{activity might do
        funky stuff}

        \subsection{Biological effects of radiation\footnote{Much of the discussion in this section follows that in
        Lilley's text \cite[chapt. 7]{l01}}}
        The biological effects of radiation have been an important topic for decades. A large
        number of studies took place after the two nuclear bomb explosions in Japan in 1945
        and after disastrous nuclear plant accidents like Chernobyl. Therefore, we have a relatively
        thorough understanding of the consequences of mass exposure to radiation. However, relatively
        little is known about the long-term effects of exposure to small doses, including subtle biological
        changes that lead to cancer and genetic defects. Here we briefly summarize the known effects.
        \subsubsection{Interactions of radiations with living tissues}
        Radiations causes damage to living tissues by depositing energy into the biological molecules.
        Depending on the type of the radiation
        and the molecules present in the tissues, the damage mechanism and effects are different.
        \paragraph{Alpha particles} An alpha particle loses energy predominantly via the interactions between its electric
        field and the electrons, in the matter it travels through. Interactions with nuclei and head-on collisions
        are quite rare. Since the speed of an alpha particle is low, it deposits more energy into the matter it
        interacts with. In tissues, $1$-MeV alpha particles travel through only a few tens of micrometers and thus
        are easily stopped by skin. However, if alpha particles are ingested and more sensitive organs are irradiated,
        serious damages can be done.
        \paragraph{Beta radiation (elections)} Electrons are much lighter and hence are easily scattered.
        Moreover, they are much more penetrating since their energy lost per distance traveled is lower.
        Finally, electrons also radiate photons
        via bremstrahlung radiations as they deccelerate. Therefore, the electrons' energy end up distributing
        across a much larger volume.
        \paragraph{Photons} Gamma rays interact with matter via photoelectric effects, Compton scattering and pair productions
        depending on th energy of the photon and the atoms it interact with.
        Since biological tissues mainly consist of elements of low atomic number, Compton
        scattering is the most important interaction. The scattered photons may interact further with the tissues
        via photoelectric effects and secondary Compton scattering. The intensity of the energy deposition falls
        off much more slowly and is hence more penetrating. Both electrons and photons can penetrate human body easily,
        although they do not deposit as much energy, they generally are more dangerous than alpha particles due to their
        penetration abilities.

@@

        Hereditary effects of radiation can occur if the reproduction cells were damaged by radiation.
        The mechanism is similar to that of delayed health effects, since DNA molecules are passed on
        to offpsrings.
        UNSCEAR estimates that risk for severe hereditary effect is about $0.3-0.5$ percent per siever to
        the first generation following the radiation exposure.\cite{u16, unscear16}
\section{Nuclear waste: classification and properties}
\label{sec:waste}
    In this section, we will first look at a classification scheme of nuclear waste and then
    examine in details the source, composition and properties of the most detrimental type, high-level waste.
    \subsection{Classification of nuclear waste}
    Different organizations have slightly different classification schemes for nuclear waste.
    However, their definitions of high-level waste (HLW), the most detrimental kind, are relatively consistent.
    The International Atomic Energy Agency (IAEA) classifies nuclear waste by considering the activity
    content and the half-life of the radionuclides.
    Activity content is a measure of how much radioactivity $\mathcal{A}(t)$, as defined in Eq.~\ref{eq:act},
    a given amount of waste contains.
    A form of waste may possesses a high activity content due to
    a high concentration of radionuclides and/or the presence of radionuclides with high activity.\cite{iaea09}
    Each form of waste, depending on the half-life, activity content, and temperature, requires different
    disposal methods. The NRC classification scheme is as follows.\cite{nrc09}

    \paragraph{High-level waste (HLW) and spent fuel}
    HLW has high activity contents and usually generates a large amount of
    heat as it radiates. The typical source of HLW is spent fuel from the nuclear reactor.
    In fact, since there is no commercial reprocessing facilities in the US (see section~\ref{sec:reproc}),
    almost all of the HLW is in the form of spent nuclear fuel from the reactor. In a nuclear reactor,
    fission of uranium-235 produces lighter radionuclides like cesium-137 and strontium-90, which
    account for most of the heat and penetrating radiation (gamma radiations). In addition to
    the fission process, some uranium-235 atoms go through radiative capture
    (see section~\ref{sec:capture}) and produce radionuclides that are
    heavier than uranium (``transuranic'') such as plutonium. The transuranic atoms do not generate
    as much heat or radiation from their subsequent decay as the fission products, but they tend to have
    a much longer half-life.
    For example, strontium-90 and cesium-137 have half-lives of about $30$ years, whereas
    the transuranic plutonium-239 has a half-life of $24,000$ years.

    \paragraph{Transuranic waste (TRU)}
    TRU is defined as materials that do not belong to high-level waste and are
    contaminated by alpha-emitting radionuclides with
    sufficiently long half-life ($>20$ years) of elements with atomic number $92$ (atomic number of uranium)
    or larger and concentration of activity larger than $3700$ Bq/g.\cite{j83,s01} The activity
    of TRU is usually low but it remains active for long period of time.

    \paragraph{Low-level waste (LLW)}
    LLW has relatively little radioactivity, produces almost no heat,
    and contains practically no
    transuranic elements. Examples are papers and tools used in cleanup operations in the reactor.
    Most LLW does not require shielding and may be buried in landfill 
    facilities after on-site short-term decay storage.
    Some LLW (for example, metal nuclear fuel cladding) may have a slightly
    higher radioactivity such that they are classified
    as intermediate-level waste (ILW). ILW are solidified in concrete and buried in shallow
    repositories.\cite{s01} However, US regulations do not make the distinction
    between LLW and ILW but different disposal methods are practiced.\cite{nrc09, s01}

    \paragraph{Uranium tailings}
    Uranium mill tailings from uranium mills are another form of waste with low activity, but
    they are not usually classified as LLW. The tailing is what is left after the uranium has been
    extracted from the ore. The decay of uranium-238 from the mill decays into
    radioactive thorium and radium and these radionuclides stay in the tailing. Notably, radium
    can decay into radon gas which is an alpha-emitter. Hence, the alpha radiation may do additional
    damage to body if the radon enters the body as one breathes.
    The uranium tailings leave the uranium mills in the form of a liquid sludge
    and can dry. Therefore, it is necessary to prevent the uranium tailings
    from contaminating water and air.

@@

    \subsection{Creation of spent fuel}
    \subsubsection{Nuclear fuel}
    For most of the commercial nuclear reactors, the energy comes from the isotope uranium-235 which
    capable of undergoing fission (see Sec.~\ref{sec:fission}). However, natural uranium consists of $99.3\%$
    non-fissile uranium-238 and only $0.7\%$ uranium-235. Therefore, for nuclear fuel, natural uranium is enriched
    via isotope separation to increase the uranium-235 content. Typical fresh nuclear fuel contains $3\%$ uranium-235
    and $97\%$ uranium-238.
    \subsubsection{Neutron absorption and radiative capture}
    \label{sec:capture}
            \includegraphics[width=0.6\textwidth]{pu.png}
            fission of uranium-235 and the radiative absorption of uranium-238.}
    Absorption of thermal neutron (with energy $\sim 0.025$ eV, the most likely speed of a free neutron at 
    a temperature of $290$ K) is the main reaction that
    occurs in the reactor. The neutrons come from a startup neutron source and subsequently from the fission
    process (see Sec.~\ref{sec:fission}).
    Since a neutron does not carry any charge, it does not experience the Coulomb repulsion potential when
    it interacts
    with a nucleus.
    Henceforth, as neutrons travels through materials, they react readily with nuclei.
    For fast neutrons, many reactions are possible but they mostly just scatter off heavy nuclei.
    But for thermal neutrons (as in the case of a nuclear reactor), neutron absorption
    is the primary mode of interaction between the nuclei and the neutrons.
    Neutron absorption is the process in which a nucleus absortb a neutron to form an unstable
    nucleus.

@@

@@

            \includegraphics[width=0.6\textwidth]{fission.png}

@@

        The reactor described by Eq.~\ref{eq:u235} is not unique in different fission products can result

@@

        On average, inside a nuclear reactor, $2.5$ neutrons are produced per fission, causing a ``chain reaction'' of subsequent

@@

        \subsubsection{Fission energy}
        We can estimate the amount of energy released in fission by considering the binding energy difference
        between the parent nuclei and the fission products. As a consequence of the masss-energy equivalence in the theory
        of special relativity, we can calculate the energy released $E$ from the atomic mass differences between the parent $m_{parent}$
        and the fission products $m_{products}$.
        \begin{equation}
            E = (m_{parent}-m_{product})c^2,
        \end{equation}
        where $c$ is the speed of light.
        Essentially, any energy carried by particles that can be slowed down 
        in the reactor can be harvested as heat to generate power.

@@

        %\subsection{reversible disposal}

>>>>

\date{\today}%
\begin{abstract}
    As nuclears around the world continuously generate highly radioactive waste,
    how we manage nuclear waste is becoming more important. In this work,
    we examine a few scientific and technical issues related to nuclear waste management and disposal.
    The principles of radioactivity and nuclear fission help us to understand the sources and properties
    of different types of nuclear waste. Focusing on the highly radioactive spent fuel, the rest
    of this work outlines the advantages and drawbacks of different spent fuel management and disposal
    methods. We study the methods used for performance assessments for Yucca Mountain Nuclear Waste Repository,
    especially in the context of volcanic disruption. Finally, we discuss some possible options for 
    nuclear waste management and disposal after
    the termination of the Yucca Mountain project.\\
    Word count: 7449\\
\end{abstract}
\maketitle
\tableofcontents
\newpage
\section{Introduction}
Nuclear power continues to play an important role in global energy supply.
The International Atomic Energy Agency (IAEA) estimates that nuclear power
consumption will increase by $40\%$ by 2030.\cite{iaea12}
Whereas the ``front end'' of the nuclear fuel cycle, such as reactor physics and power output, draws
the most attention from the public, the ``back end'' of the fuel cycle, i.e. waste management and
disposal, deserves equal, if not more attention.
Currently, the IAEA estimates that there are over 10,000 metric tons of highly radioactive nuclear waste 
generated worldwide annually,
most of which is spent nuclear fuel from nuclear reactors.\cite{iaea08, r12}
The world will have over $457,000$ metric tons of highly radioactive waste stock by the end of 2020.\cite{r12}
Highly radioactive waste remains hazardous for up to $10^6$ years after it is produced and thus
requires extremely careful disposal methods.
Nevertheless, there is currently no permanent repository in the world to dispose of these highly radioactive
waste.\footnote{The Waste Isolation Pilot Plant (WIPP) in New Mexico is designed to be a permanent radioactive
    waste repository, but only
for waste classified as Transuranic, which, by the Nuclear Regulatory Commission standard, does not have particularly
high radioactivity but only long half-life (See Sec.~\ref{sec:waste:classification}). In particular, WIPP does not accept spent nuclear
fuel.}

@@

Evidently, nuclear waste management and disposal is becoming a increasingly urgent matter globally.
In light of the high stake,
the uncertainty of the stability of a geological repository over a extremely long timescale has evolved into 
a topic not only of scientific contention but also of policy debate.
How well is the predicative power of science? How much trust should we put into the uncertainty estimates

@@

\section{Radioactivity and Its Biological Effects}
\label{sec:phys}
    Here we detail the nuclear physics underlying most of the discussions
    in this work. We will apply these results in our discussion of nuclear wastes.
    \subsection{Radioactivity\footnote{
    The derivations and in this section follows chapter $6$ in
    Krane's text \cite{k88} and chapter $1$ in Lilley's text \cite{l01}.
    }}
    Radioactive decay is the process in which most of the known detrimental radiations are generated.
    Here we review the physical principles and mathematical formalisms behind it.
        \subsubsection{Quantum theory of radioactive decays}
        The law of radioactive decay is a direct consequence of quantum mechanics.
        We can approximate the nucleus as a particle situated in a potential well with potential given by
        \begin{equation}
            \label{eq:vp}
            \tilde{V}(t) = V + V_p,
        \end{equation}
        where $V$ is the time-independent nucleus potential that gives rise to eigenstates $\ket{\psi_1}$
        and $\ket{\psi_2}$ with $\ket{\psi_2}$ having a lower (expectation value of) energy; $V_p$ is a small
        perturbation that causes the decay.
        Suppose the particle is in a state given by
        \begin{equation}
            \label{eq:state}
            \ket{\psi(t)} = c_1(t)\ket{\psi_1} + c_2(t)\ket{\psi_2},
        \end{equation}
        where $c_1,c_2$ are the coefficients of superposition.
        A decay from $\ket{\psi_1}$ to $\ket{\psi_2}$ is equivalent to
        having $|c_2(t)|^2\gg |c_1(t)|^2$ as time increases. Since we know that time-independent potentials
        only give rise to time-independent states, the fact that Eq.~\ref{eq:state} has time dependence requires
        our perturbation $V_p$ in Eq.~\ref{eq:vp} to be time dependent.

@@

        Specific activity, on the other hand, gives the activity per mass of a sample
        \begin{equation}
        \mathcal{A}_s = \frac{\mathcal{A}(0)}{{m_{sample}}}.
        \end{equation}
        Specific activity can characterize the activity level of a certain isotope or compound.
        The difference between
        activity and specific activity resembles the difference between heat capacity and specific heat
        capacity.
        The radioactive isotope used for Computer Tomography (CT) scan has a specific activity
        of $\sim 10^{8}$ Bq/g. See Table~\ref{tab:hl} for the half-lives and specific activities
        of some isotopes present in highly radioactive nuclear waste.
        \begin{table}
            \centering
            \caption{Specific activity and hal-life of isotopes commonly present in spent nuclear fuel \cite{toi}}
            \begin{ruledtabular}
                \begin{tabular}{l l l}
                Isotope & Half-life (Years) & Specific Activity (Ci/g)\\
                \hline
                Cesium-137 & $37$ & $88$\\
                Thorium-232 & $1.4 \times 10^{10}$ & $1.1\times 10^{-7}$\\
                Uranium-233 & $1.6\times 10^6$ & $9.3\times 10^{-3}$\\
                Plutonium-239 & $2.4\times 10^4$ & $0.063$\\
                Neptunium-237 & $2.1\times 10^6$ & $6.9\times 10^{-4}$\\
            \end{tabular}
            \label{tab:hl}
            \end{ruledtabular}
        \end{table}
        \subsubsection{Types of radiation}
        There are three main types of radioactive decay processes that generate radiations in nuclear waste: 
        Alpha decay, which generates alpha particles (helium-4 nuclei), Beta decay which generates beta
        particles (electrons) and gamma decay which generates gamma rays.
        The type of radiation (particle beam) generated
        have different properties and deserve separate trearments when we discuss health effects and shielding.
        \paragraph{Alpha decay}
        An alpha particle is a helium-4 atom ${}^4_2$He. In an alpha decay, a nucleus emits an alpha
        particle and decays into a different element. The emission of an alpha particle is
        particularly favorable because it does not change the parity of the nucleus and has a particularly
        high binding energy. The process can be represented by
        \begin{equation}
            {}^A_Z X_N \rightarrow {}^{A-4}_{Z-2}X'_{N-2} + \alpha.
        \end{equation}

@@

        \subsubsection{Daughter nuclei and subsequent decay\footnote{The derivation in this section follows that of
        \cite[chapt. 1.5]{l01}}}
        The resultant nuclei (called ``daughter products'') from a radioactive decay are oftern radioactive
        themselves and their decay rate can be larger or smaller than their parents. Consider a decay chain
        $A\rightarrow B \rightarrow C \rightarrow ...$ with decay constants $\lambda_A, \lambda_B,...$. Suppose
        we start with a sample of $N_0$ $A$ nuclei. The change of the number of nuclei $B$ increases as $A$
        decays and decreases s $B$ decay. Hence we have
        \begin{equation}
            \frac{dN_B}{dt} = -\lambda_B N_B + \lambda_A N_A.
        \end{equation}
        Multiply both sides by $e^{\lambda_B t}$ and substitute in $N_A=N_0 e^{-\lambda_A} t$ to get
        \begin{align*}
            \frac{dN_B}{dt}e^{\lambda_B t} + \lambda_B N_B e^{\lambda_B t} &= \lambda_A N_A e^{\lambda_B},\nonumber\\
            \frac{d}{dt} (N_B e^{\lambda_B t}) &= \lambda_A N_0 e^{(\lambda_B-\lambda_A)t}, \nonumber\\
            N_B e^{\lambda_B t} &= \lambda_A N_0 \int\limits_0^t e^{(\lambda_B - \lambda_A)u} du \nonumber\\
                                &= \frac{\lambda_A}{\lambda_B-\lambda_A}N_0 e^{(\lambda_B - \lambda_A)t}
        \end{align*}
        Hence we end up with the number of $B$ nuclei as a function of time
        \begin{equation}
            N_B(t)=\frac{\lambda_A}{\lambda_B-\lambda_A}N_0(e^{-\lambda_A t} - e^{-\lambda_B t}).
            \label{eq:daughter}
        \end{equation}
        Similar derivations can be done for the rest of the isotopes in the decay chain by considering
        one decay in a chain at a time. If the daughter isotope has a longer half-life than the parent
        ($\lambda_A \gg \lambda_B$), for small $t$, Eq.~\ref{eq:daughter} becomes
        $N_B(t)\sim N_0 e^{-\lambda_A t}$. After a long time,
        Eq.~\ref{eq:daughter} becomes $N_B(t) \sim N_0 e^{-\lambda_B t}$; the decay
        chain can be treated as the fast decay of $A$ into $B$, and then the slow decay of $B$
        into $C$.

        On the other hand, if the parent is longer lived
        ($\lambda_B \gg \lambda_A)$, after a long time, Eq.~\ref{eq:daughter} can be written as
        \begin{equation}
            N_B(t)=\frac{\lambda_A}{\lambda_B}N_0 e^{-\lambda_A t}=\frac{\lambda_A}{\lambda_B}N_A(t).
        \end{equation}
        Hence the number of nuclei $A$ and nuclei $B$ are in equilibrium determined by their decay rates, with
        $A$ dominating the nuclei count. Taking the limit as $\lambda_A \rightarrow 0$, $N_B(t)\rightarrow 0$.
        The decay is essentially the decay of $A$ directly into $C$. Hence, if we have a chain of reaction
        where each daughter nuclei has a half-life much shorter than that of its parent, then 
        isotope $A$ effectively decays into the end of the decay chain and releases a lot more radiations than
        in a single decay. Moreover, before the decay chain reaches its equilibrium, the activity of the sample
        may increase.

        \subsection{Biological effects of radiation\footnote{Much of the discussion in this section follows that in
        Lilley's text \cite[chapt. 7]{l01}}}
        The biological effects of radiation have been an important topic for decades. A large
        number of studies took place after the two nuclear bomb explosions in Japan in 1945
        and after disastrous nuclear plant accidents like Chernobyl. Therefore, we have a relatively
        thorough understanding of the consequences of mass exposure to radiation. However, relatively
        little is known about the long-term effects of exposure to small doses, including subtle biological
        changes that lead to cancer and genetic defects. Here we briefly summarize the known effects.
        \subsubsection{Interactions of radiations with living tissues}
        \label{sec:interactions}
        Radiations causes damage to living tissues by depositing energy into the biological molecules.
        Depending on the type of the radiation
        and the molecules present in the tissues, the damage mechanism and effects are different.
        \paragraph{Alpha particles} An alpha particle loses energy predominantly via the interactions between its electric
        field and the electrons, in the matter it travels through. Interactions with nuclei and head-on collisions
        are quite rare. Since the speed of an alpha particle is low, it deposits more energy into the matter it
        interacts with. In tissues, $1$-MeV alpha particles travel through only a few tens of micrometers and thus
        are easily stopped by skin. However, if alpha particles are ingested and more sensitive organs are irradiated,
        serious damages can be done.
        \paragraph{Beta radiation (elections)} Electrons are much lighter and hence are easily scattered.
        Moreover, they are much more penetrating since their energy lost per distance traveled is lower.
        Finally, electrons also radiate photons
        via bremstrahlung radiations as they deccelerate. Therefore, the electrons' energy end up distributing
        across a much larger volume.
        \paragraph{Photons} Gamma rays interact with matter via photoelectric effects, Compton scattering and pair productions
        depending on th energy of the photon and the atoms it interact with.
        Since biological tissues mainly consist of elements of low atomic number, Compton
        scattering is the most important interaction. The scattered photons may interact further with the tissues
        via photoelectric effects and secondary Compton scattering. The intensity of the energy deposition falls
        off much more slowly and is hence more penetrating. Both electrons and photons can penetrate human body easily,
        although they do not deposit as much energy, they generally are more dangerous than alpha particles due to their
        penetration abilities.

@@

        Hereditary effects of radiation can occur if the reproduction cells were damaged by radiation.
        The mechanism is similar to that of delayed health effects, since DNA molecules are passed on
        to offpsrings.
        UNSCEAR estimates that risk for severe hereditary effect is about $0.3-0.5$ percent per siever to
        the first generation following the radiation exposure.\cite{u16, unscear16}
\section{Nuclear waste: classification and properties}
\label{sec:waste}
    In this section, we will first look at a classification scheme of nuclear waste and then
    examine in details the source, composition and properties of the most detrimental type, high-level waste.
    \subsection{Classification of nuclear waste}
    \label{sec:waste:classification}
    Different organizations have slightly different classification schemes for nuclear waste.
    However, their definitions of high-level waste (HLW), the most detrimental kind, are relatively consistent.
    The International Atomic Energy Agency (IAEA) classifies nuclear waste by considering the activity
    content and the half-life of the radionuclides.
    Activity content is a measure of how much radioactivity $\mathcal{A}(t)$, as defined in Eq.~\ref{eq:act},
    a given amount of waste contains.
    A form of waste may possesses a high activity content due to
    a high concentration of radionuclides and/or the presence of radionuclides with high activity.\cite{iaea09}
    Each form of waste, depending on the half-life, activity content, and temperature, requires different
    disposal methods. The NRC classification scheme is as follows.\cite{nrc09}

    \paragraph{High-level waste (HLW) and spent fuel}
    HLW has high activity contents and usually generates a large amount of
    heat as it radiates. The typical source of HLW is spent fuel from the nuclear reactor.
    In fact, since there is no commercial reprocessing facilities in the US (see section~\ref{sec:reproc}),
    almost all of the HLW is in the form of spent nuclear fuel from the reactor.
    \paragraph{Transuranic waste (TRU)}
    TRU is defined as materials that do not belong to high-level waste and are
    contaminated by alpha-emitting radionuclides with
    sufficiently long half-life ($>20$ years) of elements with atomic number $92$ (atomic number of uranium)
    or larger and concentration of activity larger than $3700$ Bq/g.\cite{j83,s01} The activity
    of TRU is usually low but it remains active for long period of time.

    \paragraph{Low-level waste (LLW)}
    LLW has relatively little radioactivity, produces almost no heat,
    and contains practically no
    transuranic elements. Examples are papers and tools used in cleanup operations in the reactor.
    Most LLW does not require shielding and may be buried in landfill 
    facilities after on-site short-term decay storage.
    Some LLW (for example, metal nuclear fuel packaging) may have a slightly
    higher radioactivity such that they are classified
    as intermediate-level waste (ILW). ILW are solidified in concrete and buried in shallow
    repositories.\cite{s01} However, US regulations do not make the distinction
    between LLW and ILW but different disposal methods are practiced.\cite{nrc09, s01}

    \paragraph{Uranium tailings}
    Uranium mill tailings from uranium mills are another form of waste with low activity, but
    they are not usually classified as LLW.
    Over geological timescales, the stable uranium-238 decays into radioactive thorium and radium 
    These radionuclides remain after the uraninium (all of its isotopes)
    has been extracted from the ore. Notably, radium
    can decay into radon gas which is an alpha-emitter. Hence, the alpha radiation may do additional
    damage to body if the radon enters the body as one breathes.
    The uranium tailings leave the uranium mills in the form of a liquid sludge
    and can dry. Therefore, it is necessary to prevent the uranium tailings
    from contaminating water and air.

@@

    \subsection{Creation of spent fuel}
    \label{sec:waste:spentfuel}
    \subsubsection{Nuclear fuel}
    For most of the commercial nuclear reactors, the energy comes from the isotope uranium-235 which
    capable of undergoing fission (see Sec.~\ref{sec:fission}). However, natural uranium consists of $99.3\%$
    non-fissile uranium-238 and only $0.7\%$ uranium-235. Therefore, for nuclear fuel, natural uranium is enriched
    via isotope separation to increase the uranium-235 content. Typical fresh nuclear fuel contains $3\%$ uranium-235
    and $97\%$ uranium-238.
    \subsubsection{Neutron absorption and radiative capture}
    \label{sec:capture}
    Absorption of thermal neutron (with energy $\sim 0.025$ eV, the most likely speed of a free neutron at 
    a temperature of $290$ K) is the main reaction that
    occurs in the reactor. The neutrons come from a startup neutron source and subsequently from the fission
    process (see Sec.~\ref{sec:fission}).
    Since a neutron does not carry any charge, it does not experience the Coulomb repulsion potential when
    it interacts
    with a nucleus.
    Henceforth, as neutrons travels through materials, they react readily with nuclei.
    For fast neutrons, many reactions are possible but they mostly just scatter off heavy nuclei.
    But for thermal neutrons (as in the case of a nuclear reactor), neutron absorption
    is the primary mode of interaction between the nuclei and the neutrons.
    Neutron absorption is the process in which a nucleus absortb a neutron to form an unstable
    nucleus.

@@

            \includegraphics[width=0.8\textwidth]{pu.png}
            fission of uranium-235 and the radiative absorption of uranium-238. Adapted from \cite{iync_2013}}

@@

@@

        The reaction described by Eq.~\ref{eq:u235} is not unique because different fission products can result

@@

        On average, inside a nuclear reactor, $2.5$ neutrons are produced per fission. These neutrons
        continue to induce fissions of other uranium-235 nuclei, causing a ``chain reaction'' of subsequent

@@

            \includegraphics[width=0.8\textwidth]{fission.png}
        \subsubsection{Fission energy}
        We can estimate the amount of energy released in fission by considering the binding energy difference
        between the parent nuclei and the fission products. As a consequence of the masss-energy equivalence in the theory
        of special relativity, we can calculate the energy released $E$ from the atomic mass differences between the parent $m_{parent}$
        and the fission products $m_{products}$.
            E = (m_{parent}-m_{product})c^2,
        where $c$ is the speed of light.
        Essentially, any energy carried by particles that can be slowed down 
        in the reactor can be harvested as heat to generate power.

@@

\section{Discussions}
Nuclear waste management and disposal is certainly becoming an increasingly pressing issue.
We explore the creation, management and disposal of highly radioactive nuclear waste and the physical
princples underlying those processes. The increasingly overpacked spent fuel pools and the challenges
associated with geological repositories are the weak links in this chain.
